Notes on generating initial models for mapping into {\tt MAESTRO}

### 2010-09-03 docs/initial_models/initial_models.tex
remove stray 'g' factors that Andy pointed out
some updates

MAESTRO needs to be given model data with the same resolution as the base state.
(Note that for spherical problems we set 
$\Delta r = \Delta x/\mathtt{drdxfac}$.)  
We need to generate this from the raw data,
 $\rho^{\raw}, T^{\raw}, p^{\raw}$, and
$X^{\raw}$.  Here is the raw data file for each test problem:
\begin{eqnarray}
{\tt test2} & \rightarrow & \mathrm{none-it~ is~ generated~ analytically} \nonumber \nonumber \\
{\tt test\_convect} & \rightarrow & \mathrm{none-it~ is~ generated~ analytically} \nonumber \nonumber \\
{\tt wdconvect} & \rightarrow & {\tt initial\_models/kepler/model\_6.e8.raw} \nonumber \\
{\tt spherical\_heat} & \rightarrow & \mathrm{none-it~ is~ generated~ analytically} \nonumber \\
{\tt xrb} & \rightarrow & {\tt initial\_models/xrb/sorted\_xrb.raw} \nonumber
\end{eqnarray}

We use a fortran subroutine
to interpolate the raw data, yielding the model data, $\rho^{\model},
T^{\model}, p^{\model}$, and $X^{\model}$.  The fortran subroutine
then uses an iterative procedure to modify the model data so that it
is thermodynamically consistent with the {\tt MAESTRO} equation of
state (EOS), and also satisfies our chosen hydrostatic equilibrium
(HSE) discretization,
\begin{equation}
\frac{p_r - p_{r-1}}{\Delta r} = \frac{\rho_r + \rho_{r-1}}{2}g,\label{HSE Discretization}
\end{equation}
to a user defined tolerance.  Here are the fortran subroutines for each test problem:
\begin{eqnarray}
{\tt test2} & \rightarrow & {\tt initial\_models/test2/init\_1d.f90} \nonumber \\
{\tt test\_convect} & \rightarrow & {\tt initial\_models/test2/init\_1d.f90} \nonumber \\
{\tt wdconvect} & \rightarrow & {\tt initial\_models/kepler/init\_1d.f90} \nonumber \\
{\tt spherical\_heat} & \rightarrow & {\tt initial\_models/spherical/init\_1d.f90} \nonumber \\
{\tt xrb} & \rightarrow & {\tt initial\_models/xrb/init\_1d.f90} \nonumber
\end{eqnarray}
The model data is not generated at run-time - it must be generated in
advance of running any MAESTRO examples.  The inputs file should point
to the file containing the model data.

\subsection{Multilevel Plane-Parallel}
We do the same thing as the single-level case except that we generate
model data with the same resolution as the finest base state
resolution.

\section{Creating the Initial Data from the Model Data}
In {\tt base\_state.f90}, using the subroutine {\tt
  init\_base\_state} (which is actually a terrible name since we
are setting the initial data, not the base state), we set the initial
data $\rho^{\init}, T^{\init}, p^{\init}$ and $X^{\init}$ equal to the
model data.  Then, we set $h^{\init} =
h(\rho^{\init},T^{\init},X^{\init})$ through the EOS.  Note that in the code,
$p^{\init}$ is also overwritten, but the value of $p^{\init}$ does not
change since we called $p^{\model} =
p^{\model}(\rho^{\model},T^{\model},X^{\model})$ at the end of the
initial model generation routines in \S~\ref{Sec:Creating the Model Data from Raw
  Data}.

\subsection{Multilevel Plane-Parallel}
We do the same thing as the single-level case except that we generate
initial data at the non-finest levels by using linear interpolation
from the two nearest model points (see Figure \ref{Fig:Multilevel
  Initialization}).  Note that the initial data is not in HSE nor is
it thermodynamically consistent on non-finest cells.  We will enforce
these on the base state and full state later.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tpb]
\centering
\includegraphics[width=0.4in]{\figpath/multilevel_init}\hspace{0.2in}
\begin{minipage}[b]{5.0in}
\caption[Multi-level initial model initialization]
 {Multilevel Initialization.  The data from the initial model
  is represented by the dots on the right.  The initial data at the
  finest level is represented by the circles.  The initial data at
  non-finest levels is represented by the squares.  We copy data from
  the dots directly into the circles.  We use linear interpolation
  using the two nearest points to copy data from the dots to the
  squares.\vspace{2em}}
\end{minipage}
\label{Fig:Multilevel Initialization}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Next we integrate up from the coarse-fine interface to the fine-cell center:
\begin{equation}
\frac{p_r^l - p_{r-\myhalf}^l}{\Delta r^l/2} = 
  \frac{\rho_r^l + \rho_{r-\myhalf}^l}{2} \,
  \frac{g_r^l + g_{r-\myhalf}^l}{2}
\end{equation}
We can rewrite this as an expression for the pressure at the fine-cell center:
\begin{equation}
p_r^l = p_{r-\myhalf}^l + \frac{\Delta r^l g}{8}
  \left(\rho_r^l + \rho_{r-\myhalf}^l\right)
  \left(g_r^l + g_{r-\myhalf}^l\right).
  \label{eq:itof}
\end{equation}
Combining equations \ref{eq:ctoi} and \ref{eq:itof} gives
\begin{eqnarray}
p_r^l = p_{\sfrac{r}{2}-1}^{l-1} &+& 
     \frac{\Delta r^{l-1}}{8} \left(\rho_{r-\myhalf}^l + \rho_{\sfrac{r}{2}-1}^{l-1}\right)
                                \left(   g_{r-\myhalf}^l +    g_{\sfrac{r}{2}-1}^{l-1}\right) \nonumber \\
 &+& \frac{\Delta r^l}{8} \left(\rho_r^l + \rho_{r-\myhalf}^l\right)
                            \left(   g_r^l +    g_{r-\myhalf}^l\right).
\end{eqnarray}
We can simplify using
\begin{equation}
\Delta r^{l-1} = 2\Delta r^l,
\end{equation}
and by interpolating the cell-centered densities to the coarse-fine interface as:
\begin{equation}
\rho_{r-\myhalf}^l = \frac{2}{3}\rho_r^l + \frac{1}{3}\rho_{\sfrac{r}{2}-1}^{l-1}.
\end{equation}
Because we carry both the cell- and edge-centered gravitational accelerations, we
do not need to interpolate $g$ to the interface.
Simplifying, we have
\begin{eqnarray}
p_r^l = p_{\sfrac{r}{2}-1}^{l-1} &+& 
   \frac{\Delta r^l}{4}\left(\frac{2}{3}\rho_r^l + 
                             \frac{4}{3}\rho_{\sfrac{r}{2}-1}^{l-1} \right)
                       \left(   g_{r-\myhalf}^l +    g_{\sfrac{r}{2}-1}^{l-1}\right) \nonumber \\
  &+& \frac{\Delta r^l g}{8}\left(\frac{5}{3}\rho_r^l + 
                                  \frac{1}{3}\rho_{\sfrac{r}{2}-1}^{l-1}\right) 
                       \left(   g_r^l +    g_{r-\myhalf}^l\right) \enskip .
\end{eqnarray}
Finally, we note for constant $g$, this simplifies to:
\begin{equation}
p_r^l = p_{\sfrac{r}{2}-1}^{l-1} + 
  \frac{3\Delta r^l g}{4}\left(\rho_{\sfrac{r}{2}-1}^{l-1} + \rho_r^l\right).\label{Coarse-Fine Stencil}
\end{equation}

When integrating across a fine-coarse interface (see Figure~\ref{fig:ftoc}), the proceduce is similar.
\begin{figure}[t]
\centering
\includegraphics[width=2.5in,angle=90]{\figpath/ftoc}
\caption{\label{fig:ftoc} A fine-coarse interface in the 1-d base state}
\end{figure}
The expression for general gravity becomes:
\begin{eqnarray}
p_{(r+1)/2}^{l-1} = p_{r}^l &+&
   \frac{\Delta r^l}{4}\left(\frac{2}{3}\rho_r^l + 
                             \frac{4}{3}\rho_{(r+1)/2}^{l-1} \right)
                       \left(   g_{(r+1)/2 -\myhalf}^{l-1} +    g_{(r+1)/2}^{l-1}\right) \nonumber \\
  &+& \frac{\Delta r^l g}{8}\left(\frac{5}{3}\rho_r^l + 
                                  \frac{1}{3}\rho_{(r+1)/2}^{l-1}\right) 
                       \left(   g_r^l +    g_{(r+1)/2 -\myhalf}^{l-1} \right) \enskip .
\end{eqnarray}
and for spatially-constant gravity, it simplifies to:
\begin{equation}
p_{(r+1)/2}^{l-1} = p_{r}^l + \frac{3\Delta r^l g}{4}\left(\rho_{r}^l+\rho_{(r+1)/2}^{l-1}\right).\label{Fine-Coarse Stencil}
\end{equation}

>>>>

We have found that for the best performance, the MAESTRO initialization procedure should be given model data with the same resolution as the base state resolution, $\Delta r$.  For planar problems, $\Delta r = \Delta x$.  For multilevel planar problems, we use $\Delta r$ equal to $\Delta x$ at the finest resolution.  For spherical problems we set $\Delta r = \Delta x/\mathtt{drdxfac}$.  We generate our initial model either analytically or from raw data,
 $\rho^{\raw}, T^{\raw}, p^{\raw}$, and $X^{\raw}$.  
Here is the raw data file for each test problem:
\begin{eqnarray}
{\tt test2} & \rightarrow & \mathrm{none-it~ is~ generated~ analytically} \nonumber \nonumber \\
{\tt test\_convect} & \rightarrow & \mathrm{none-it~ is~ generated~ analytically} \nonumber \nonumber \\
{\tt wdconvect} & \rightarrow & {\tt initial\_models/kepler/model\_6.e8.raw} \nonumber \\
{\tt spherical\_heat} & \rightarrow & \mathrm{none-it~ is~ generated~ analytically} \nonumber \\
{\tt xrb} & \rightarrow & {\tt initial\_models/xrb/sorted\_xrb.raw} \nonumber
\end{eqnarray}

We use a fortran subroutine
to interpolate the raw data, yielding the model data, $\rho^{\model},
T^{\model}, p^{\model}$, and $X^{\model}$.  The fortran subroutine
then uses an iterative procedure to modify the model data so that it
is thermodynamically consistent with the {\tt MAESTRO} equation of
state (EOS), and also satisfies our chosen hydrostatic equilibrium
(HSE) discretization,
\begin{equation}
\frac{p_r - p_{r-1}}{\Delta r} = \frac{\rho_r + \rho_{r-1}}{2}g_{r-\myhalf},\label{HSE Discretization}
\end{equation}
to a user defined tolerance.  Here are the fortran subroutines for each test problem:
\begin{eqnarray}
{\tt test2} & \rightarrow & {\tt initial\_models/test2/init\_1d.f90} \nonumber \\
{\tt test\_convect} & \rightarrow & {\tt initial\_models/test2/init\_1d.f90} \nonumber \\
{\tt wdconvect} & \rightarrow & {\tt initial\_models/kepler/init\_1d.f90} \nonumber \\
{\tt spherical\_heat} & \rightarrow & {\tt initial\_models/spherical/init\_1d.f90} \nonumber \\
{\tt xrb} & \rightarrow & {\tt initial\_models/xrb/init\_1d.f90} \nonumber
\end{eqnarray}
The model data is not generated at run-time - it must be generated in
advance of running any MAESTRO examples.  The inputs file should point
to the file containing the model data.

\section{Creating the Initial Data from the Model Data}
In {\tt base\_state.f90}, using the subroutine {\tt
  init\_base\_state} (which is actually a terrible name since we
are computing 1D initial data, which is not quite the same thing 
as the base satte), we set the initiail
data $\rho^{\init}, T^{\init}, p^{\init}$ and $X^{\init}$ equal to the
model data.  Then, we set $p^{\init},h^{\init} =
p,h(\rho^{\init},T^{\init},X^{\init})$ through the EOS.  Note that 
we overwrite the existing value of $p^{\init}$ but this should not change
the value since we called 
$p^{\model} = p^{\model}(\rho^{\model},T^{\model},X^{\model})$ at the end of the
initial model generation routines in 
\S~\ref{Sec:Creating the Model Data from Raw Data}.

For multilevel planar problems, we generate
initial data at the non-finest levels by using linear interpolation
from the two nearest model points (see Figure \ref{Fig:Multilevel
  Initialization}).  Note that the initial data is not in HSE nor is
it thermodynamically consistent on non-finest cells.  We will enforce
these on the base state and full state later.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tpb]
\centering
\includegraphics[width=0.4in]{\figpath/multilevel_init}\hspace{0.2in}
\begin{minipage}[b]{5.0in}
\caption[Multi-level initial model initialization]
 {Multilevel Initialization.  The data from the initial model
  is represented by the dots on the right.  The initial data at the
  finest level is represented by the circles.  The initial data at
  non-finest levels is represented by the squares.  We copy data from
  the dots directly into the circles.  We use linear interpolation
  using the two nearest points to copy data from the dots to the
  squares.\vspace{2em}}
\end{minipage}
\label{Fig:Multilevel Initialization}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Next we integrate up from the coarse-fine interface to the fine-cell center:
\begin{equation}
\frac{p_r^l - p_{r-\myhalf}^l}{\Delta r^l/2} = 
  \frac{\rho_r^l + \rho_{r-\myhalf}^l}{2} \,
  \frac{g_r^l + g_{r-\myhalf}^l}{2}
\end{equation}
We can rewrite this as an expression for the pressure at the fine-cell center:
\begin{equation}
p_r^l = p_{r-\myhalf}^l + \frac{\Delta r^l}{8}
  \left(\rho_r^l + \rho_{r-\myhalf}^l\right)
  \left(g_r^l + g_{r-\myhalf}^l\right).
  \label{eq:itof}
\end{equation}
Combining equations \ref{eq:ctoi} and \ref{eq:itof} gives
\begin{eqnarray}
p_r^l = p_{\sfrac{r}{2}-1}^{l-1} &+& 
     \frac{\Delta r^{l-1}}{8} \left(\rho_{r-\myhalf}^l + \rho_{\sfrac{r}{2}-1}^{l-1}\right)
                                \left(   g_{r-\myhalf}^l +    g_{\sfrac{r}{2}-1}^{l-1}\right) \nonumber \\
 &+& \frac{\Delta r^l}{8} \left(\rho_r^l + \rho_{r-\myhalf}^l\right)
                            \left(   g_r^l +    g_{r-\myhalf}^l\right).
\end{eqnarray}
We can simplify using
\begin{equation}
\Delta r^{l-1} = 2\Delta r^l,
\end{equation}
and by interpolating the cell-centered densities to the coarse-fine interface as:
\begin{equation}
\rho_{r-\myhalf}^l = \frac{2}{3}\rho_r^l + \frac{1}{3}\rho_{\sfrac{r}{2}-1}^{l-1}.
\end{equation}
Because we carry both the cell- and edge-centered gravitational accelerations, we
do not need to interpolate $g$ to the interface.
Simplifying, we have
\begin{eqnarray}
p_r^l = p_{\sfrac{r}{2}-1}^{l-1} &+& 
   \frac{\Delta r^l}{4}\left(\frac{2}{3}\rho_r^l + 
                             \frac{4}{3}\rho_{\sfrac{r}{2}-1}^{l-1} \right)
                       \left(   g_{r-\myhalf}^l +    g_{\sfrac{r}{2}-1}^{l-1}\right) \nonumber \\
  &+& \frac{\Delta r^l}{8}\left(\frac{5}{3}\rho_r^l + 
                                  \frac{1}{3}\rho_{\sfrac{r}{2}-1}^{l-1}\right) 
                       \left(   g_r^l +    g_{r-\myhalf}^l\right) \enskip .
\end{eqnarray}
Finally, we note for constant $g$, this simplifies to:
\begin{equation}
p_r^l = p_{\sfrac{r}{2}-1}^{l-1} + 
  \frac{3\Delta r^l g}{4}\left(\rho_{\sfrac{r}{2}-1}^{l-1} + \rho_r^l\right).\label{Coarse-Fine Stencil}
\end{equation}

When integrating across a fine-coarse interface (see Figure~\ref{fig:ftoc}), the proceduce is similar.
\begin{figure}[t]
\centering
\includegraphics[width=2.5in,angle=90]{\figpath/ftoc}
\caption{\label{fig:ftoc} A fine-coarse interface in the 1-d base state}
\end{figure}
The expression for general gravity becomes:
\begin{eqnarray}
p_{(r+1)/2}^{l-1} = p_{r}^l &+&
   \frac{\Delta r^l}{4}\left(\frac{2}{3}\rho_r^l + 
                             \frac{4}{3}\rho_{(r+1)/2}^{l-1} \right)
                       \left(   g_{(r+1)/2 -\myhalf}^{l-1} +    g_{(r+1)/2}^{l-1}\right) \nonumber \\
  &+& \frac{\Delta r^l}{8}\left(\frac{5}{3}\rho_r^l + 
                                  \frac{1}{3}\rho_{(r+1)/2}^{l-1}\right) 
                       \left(   g_r^l +    g_{(r+1)/2 -\myhalf}^{l-1} \right) \enskip .
\end{eqnarray}
and for spatially-constant gravity, it simplifies to:
\begin{equation}
p_{(r+1)/2}^{l-1} = p_{r}^l + \frac{3\Delta r^l g}{4}\left(\rho_{r}^l+\rho_{(r+1)/2}^{l-1}\right).\label{Fine-Coarse Stencil}
\end{equation}
